A Mechanised Library for Generic (Co)Programming

### 2024-02-23 paper/paper.tex
minor
Fixed problem in paper
Minimal, just started working on Section 3 and correct David's obnoxious uncompilable code.

First we prove that the hylo equation has a unique solution
\[
  x = a \comp F(x) \comp c
\]
for every recursive coalgebra $c$ for a polynomial $F$. Then we prove each
recursion scheme by reducing it to a hylo. The following recursion schemes can
all be proven by reducing them to an hylo: folds, mutual recursive functions,
recursion schemes from comonads and their respective coinductive dual.

@@

\section{Mechanising Extractable Container Functors}

\dcas{Subsections are tentative topics to remember what I need to fill in here,
but these are not final}

Our mechanisation works in the category of Coq types, and our goals are: (1)
the mechanisation of hylomorphisms and their laws should be axiom-free; and (2)
Coq hylomorphisms should be extracted to recursive OCaml functions that use no
instances of \mintinline{OCaml}{Obj.magic}.  Although the use of
\mintinline{OCaml}{Obj.magic} is safe in the code extracted by Coq, it has been
proven that, for higher-order programs, simple interoperations can lead to
incorrect behaviour or even segfaults~\cite{forster:hal-04329663}.  Therefore,
the avoidance of axioms and  \mintinline{OCaml}{Obj.magic} means that the
extracted code will have stronger safety guarantees, and will allow the safe
interoperation with other OCaml code.

\subsection{Functors and Containers}

In Coq, to construct least/greatest fixpoints of functors, these need to build
\emph{strictly positive types}. A common approach to construct such strictly
positive endofunctors is to use \emph{container functors}~\cite{AbbottAG05}.
However, reasoning about container and container functor equality requires to
consider: (a) functional extensionality; (b) heterogeneous equality. This is
problematic, due to our goals to remain axiom-free. Furthermore, standard
representations of containers in Coq will lead to OCaml code that contains
\mintinline{OCaml}{Obj.magic}. We will explain why, and our mechanisation of
containers in the remainder of this section.

@@

\subsubsection{Polynomial Functors as Container Extensions}

Polynomial functors can be defined as container extensions~\ref{fig:poly}. We
omit here the proofs that these container extensions are isomorphic to
polynomials, but this proof is straightforward. To illustrate how the identity,
constant, sum and product container extensions work, consider the
tree-generating functor (i.e.\ the functor whose fixpoint is a binary tree):
$F\; X = \mathbb{N} + X \times X$.

@@

\subsubsection{Equality of Container Extensions}

The equality of container extensions needs to be extensional, since the
contents are obtained by applying a function. We want to equate container
extensions that have the same shape, and that, for equal shape and position,
they return the same element. To avoid the use of functional extensionality
axioms, we capture this relation with the following inductive proposition in
Coq:
\begin{minted}{Coq}
Inductive AppR `{F : Cont Sh P} {X} (x y : App F X) : Prop :=
| AppR_ext (Es : shape x = shape y)
           (Ek : forall e1 e2, val e1 = val e2 -> cont x e1 = cont y e2).
\end{minted}
Note that  we do not care about the validity proof of the positions, only their
value. This is to simplify (slightly) our proofs. This relation is
trivially \emph{reflexive}, \emph{transitive}, and \emph{symmetric}.

@@

\subsection{Types and Morphisms}

We address the different forms of equality by coupling every type with an
associated equivalence relation, and every morphism with a proof that it is
\emph{respectful} (i.e. \emph{proper} morphisms with respect to the function
\emph{respectfulness} relation). We use \mintinline{Coq}{=e} to denote the
equivalence relation associated with a type. Every Coq type is associated by
default with the standard Coq propositional equality, unless a different
equivalence is specified. We use the type-class mechanism to help Coq's code
extraction mechanism to remove any occurence of custom equivalence relations
in the extracted OCaml code.
\begin{minted}{Coq}
Reserved Notation "f =e g" (at level 70, no associativity).
Class equiv A : Type :=
  MkEquiv
    { eqRel : A -> A -> Prop;
      e_refl : forall x, eqRel x x;
      e_sym : forall x y, eqRel x y -> eqRel y x;
      e_trans : forall x y z, eqRel x y -> eqRel y z -> eqRel x z;
    }.
Notation "f =e g" := (eqRel f g).
\end{minted}
Given types \mintinline{Coq}{A} and  \mintinline{Coq}{B}, with their respective
equivalence relations \mintinline{Coq}{eA : A -> A -> Prop} and
\mintinline{Coq}{eB : B -> B -> Prop}, the we define the type
\mintinline{Coq}{A ~> B} to represents \emph{proper} morphisms of the
respectfulness relation of \mintinline{Coq}{R} to \mintinline{Coq}{R'}.
\begin{minted}{Coq}
Structure morph A {eA : equiv A} B {eB : equiv B} :=
  MkMorph
    { app :> A -> B;
      app_eq : forall x y, x =e y -> app x =e app y
    }.
Notation "A ~> B" := (@morph A _ B _).
\end{minted}
Note that we leverage on Coq's type class mechanism to fill in the necessary
equivalence relations. Since Coq's code extraction mechanism will erase any
occurrence of \mintinline{Coq}{Prop} in the code, so objects of type
\mintinline{Coq}{A ~> B} will be extracted to the OCaml equivalent to
\mintinline{Coq}{A -> B}.

\subsubsection{Polynomial Types}

We define a number of equivalences for polynomial types. These equivalences are
be used when building proper morphisms that manipulate their respective types.
\begin{minted}{Coq}
Instance ext_eq (A : Type) `{eq_B : equiv B} : equiv (A -> B).
Instance pair_eq `{eq_A : equiv A} `{eq_B : equiv B} : equiv (A * B).
Instance sum_eq `{eq_A : equiv A} `{eq_B : equiv B} : equiv (A + B).
Instance prop_eq : equiv Prop.
Instance pred_sub `{eA : equiv A} {P : A -> Prop} : equiv {a : A | P a}.
\end{minted}

>>>>

First we prove that the hylo equation has a unique solution
\[
  x = a \comp F(x) \comp c
\]
for every recursive coalgebra $c$ for a polynomial $F$. Then we prove each
recursion scheme by reducing it to a hylo. The following recursion schemes can
all be proven by reducing them to an hylo: folds, mutual recursive functions,
recursion schemes from comonads and their respective coinductive dual.
\mpav{Moved this from Section 3}
goals are: (1) the
mechanisation of hylomorphisms and their laws should be axiom-free; and (2) Coq
hylomorphisms should be extracted to recursive OCaml functions that use no
instances of \mintinline{OCaml}{Obj.magic}.  Although the use of
\mintinline{OCaml}{Obj.magic} is safe in the code extracted by Coq, it has been
proven that, for higher-order programs, simple interoperations can lead to
incorrect behaviour or even segfaults~\cite{forster:hal-04329663}.  Therefore,
the avoidance of axioms and  \mintinline{OCaml}{Obj.magic} means that the
extracted code will have stronger safety guarantees, and will allow the safe
interoperation with other OCaml code.

@@

\section{Mechanising Extractable Container Functors}
\dcas{Subsections are tentative topics to remember what I need to fill in here,
but these are not final}

In order to abstract recursion patters we need to be able to talk about functors
which have the suitable fixed-point properties. Instances of these functors are
the polynomial ones which are in one-to-one correspondence with the so called
\emph{containers}~\cite{AbbottAG05}.
In this section we introduce containers

\subsection{Functors and Containers}
\label{sec:containers}
In order to To construct fixpoints of functors, these need to build
\emph{strictly positive types}. A common approach to construct such strictly
positive endofunctors is to use \emph{container functors}~\cite{AbbottAG05}.
However, reasoning about container and container functor equality requires to
consider both functional extensionality and heterogeneous equality. This is
problematic since, as stated in Section~\ref{sec:intro}, our goals to remain
axiom-free.  Furthermore, standard representations of containers in Coq will
lead to OCaml code that contains \mintinline{OCaml}{Obj.magic}. We will explain
why, and our mechanisation of containers in the remainder of this section.

@@

\subsubsection{Polynomial Functors as Container Extensions}
Polynomial functors can be defined as container extensions~\ref{fig:poly}. We
omit here the proofs that these container extensions are isomorphic to
polynomials, but this proof is straightforward. To illustrate how the identity,
constant, sum and product container extensions work, consider the
tree-generating functor (i.e.\ the functor whose fixpoint is a binary tree):
$F\; X = \mathbb{N} + X \times X$.

@@

\subsubsection{Equality of Container Extensions}
The equality of container extensions needs to be extensional, since the
contents are obtained by applying a function. We want to equate container
extensions that have the same shape, and that, for equal shape and position,
they return the same element. To avoid the use of functional extensionality
axioms, we capture this relation with the following inductive proposition in
Coq:
\begin{minted}{Coq}
Inductive AppR `{F : Cont Sh P} {X} (x y : App F X) : Prop :=
| AppR_ext (Es : shape x = shape y)
           (Ek : forall e1 e2, val e1 = val e2 -> cont x e1 = cont y e2).
\end{minted}
Note that  we do not care about the validity proof of the positions, only their
value. This is to simplify (slightly) our proofs. This relation is
trivially \emph{reflexive}, \emph{transitive}, and \emph{symmetric}.

@@

\subsection{Types and Morphisms}
\label{sec:types-morphisms}
We address the different forms of equality by coupling every type with an
associated equivalence relation, and every morphism with a proof that it is
\emph{respectful} (i.e. \emph{proper} morphisms with respect to the function
\emph{respectfulness} relation). We use \mintinline{Coq}{=e} to denote the
equivalence relation associated with a type. Every Coq type is associated by
default with the standard Coq propositional equality, unless a different
equivalence is specified. We use the type-class mechanism to help Coq's code
extraction mechanism to remove any occurence of custom equivalence relations
in the extracted OCaml code.
\begin{minted}{Coq}
Reserved Notation "f =e g" (at level 70, no associativity).
Class equiv A : Type :=
  MkEquiv
    { eqRel : A -> A -> Prop;
      e_refl : forall x, eqRel x x;
      e_sym : forall x y, eqRel x y -> eqRel y x;
      e_trans : forall x y z, eqRel x y -> eqRel y z -> eqRel x z;
    }.
Notation "f =e g" := (eqRel f g).
\end{minted}
Given types \mintinline{Coq}{A} and  \mintinline{Coq}{B}, with their respective
equivalence relations \mintinline{Coq}{eA : A -> A -> Prop} and
\mintinline{Coq}{eB : B -> B -> Prop}, the we define the type
\mintinline{Coq}{A ~> B} to represents \emph{proper} morphisms of the
respectfulness relation of \mintinline{Coq}{R} to \mintinline{Coq}{R'}.
\begin{minted}{Coq}
Structure morph A {eA : equiv A} B {eB : equiv B} :=
  MkMorph
    { app :> A -> B;
      app_eq : forall x y, x =e y -> app x =e app y
    }.
Notation "A ~> B" := (@morph A _ B _).
\end{minted}
Note that we leverage on Coq's type class mechanism to fill in the necessary
equivalence relations. Since Coq's code extraction mechanism will erase any
occurrence of \mintinline{Coq}{Prop} in the code, so objects of type
\mintinline{Coq}{A ~> B} will be extracted to the OCaml equivalent to
\mintinline{Coq}{A -> B}.

\subsubsection{Polynomial Types}
We define a number of equivalences for polynomial types. These equivalences are
be used when building proper morphisms that manipulate their respective types.
\begin{minted}{Coq}
Instance ext_eq (A : Type) `{eq_B : equiv B} : equiv (A -> B).
Instance pair_eq `{eq_A : equiv A} `{eq_B : equiv B} : equiv (A * B).
Instance sum_eq `{eq_A : equiv A} `{eq_B : equiv B} : equiv (A + B).
Instance prop_eq : equiv Prop.
Instance pred_sub `{eA : equiv A} {P : A -> Prop} : equiv {a : A | P a}.
\end{minted}
